Owlifier: Creating OWL-DL Ontologies from Simple Spreadsheet-Based Knowledge Descriptions\tnoteref{t1}

### 2009-08-29 docs/pubs/ISEI6/main.tex
final version

Efforts to engage scientists in the development of ontologies typically leverage the W3C Web Ontology Language (OWL) \citep{smith04:_owl_web_ontol_languag_guide} as a standard XML syntax for representing and sharing ontologies. A key advantage of OWL is that it is supported by a wide range of generic tools, including editors \citep{knublauch04:_editin_descr_logic_ontol_with,kalyanpur05:_swoop}, reasoning systems \citep{sirin07:_pellet,tsarkov06:_fact_descr_logic_reason}, query languages \citep{prudhommeaux08:_sparq_query_languag_for_rdf,motik05:_query_answer_for_owl_dl_with_rules}, and storage technologies \citep{carroll04:_jena,broekstra02:_sesam}. However, most of these tools are primarily targeted at experts in knowledge engineering and software development familiar with the underlying description logic semantics of OWL-DL \citep{grau08:_owl}. This is especially true with ontology editors (such as Protege, SWOOP, etc.), which allow for very detailed ontology specifications, but at the same time require considerable understanding of the underlying ontology formalisms and syntax.  A well-funded ontology development effort, the Gene Ontology (GO), has been underway for over a decade within the molecular biology community, and has had success in enriching genomic and biomedical data resources with annotations to community-based ontologies. A major lesson learned from these efforts was the importance of keeping the ontology focused, and indeed GO has a highly delimited scope \citep{bada04:_short_study_succes_of_gene_ontol}. Ecology, on the other hand, encompasses an extremely broad range of scales and disciplines, such that a coordinated ontology development effort would need to involve a large number of experts to cover the diverse types of information relevant to ecological investigations.  At this point, we believe that ecological researchers need to begin constructing ontologies for their own specialized sub-disciplines, and due to the substantial expertise in formal knowledge representation required in most ontology editing tools, new tools are needed to enable wide-scale creation and adoption of ontologies in ecology.

This paper presents a novel approach for ontology creation that aims at being more intuitive for ecologists and that can be used to rapidly construct large ontologies of ecological terms (classes and properties), which can then be used in a subsequent step for annotation of scientific data (instances) \citep{bowers08:_concep_model_framew_for_expres,berkley09:_improv_data_discov_metad_repos_seman_searc}.  Our approach is to allow scientists to use common and familiar spreadsheet-based tools (e.g., Microsoft Excel, Apple Numbers, Open Office Spreadsheet) to describe, in an intuitive way, different aspects of an ontology, and then to take these descriptions and convert them into full-fledged OWL ontologies using a software application called \owlifier.

An \owlifier\ spreadsheet consists of a set of \emph{blocks} that have a predefined template structure for users to fill in. Each non-empty row in an \owlifier\ table constitutes a block. Each block defines different aspects of an ontology including classes, subclasses, synonyms, and properties.  We also provide blocks for plain-text descriptions of classes and properties, and for referencing one or more existing ontologies (e.g., to extend an existing ontology or to define ontology articulations). Blocks can be sparse (inheriting from previous blocks), which can further simplify the creation of large ontologies. We omit several OWL features associated with properties (e.g., role inclusion axioms, domain and range property restrictions, and data-type properties among others) as these can be confusing to construct for non-experts and are more suitable for experienced knowledge modelers.

While not as expressive as OWL-DL, our approach can produce ontology structures that are essential for improved data discovery and integration \citep{madin07:_ontol_for_descr_and_synth}. Just as important, because spreadsheet tools are already frequently used by ecologists to store and analyze data, \owlifier\ can provide ecologists with a familiar and accessible user interface for ontology creation. This approach also leverages the easy-to-use interfaces provided by many spreadsheet tools for organizing and manipulating tabular data, e.g., via cut/copy/paste, search/replace, sort, split-window, track-changes, freeze panes, and so on.  In this way, an ecologist can easily construct (or load) a set of terms, and then incrementally organize these into class hierarchies, properties, and constraints.  The use of an editing environment that is familiar to scientists can significantly help improve the speed and understanding of ontology construction and avoid the often time-consuming task of locating, downloading, installing, and learning fundamentally new software applications and interfaces (e.g., CMAP-COE \citep{hayes05:_collab_knowl_captur_in_ontol}, Protege \citep{knublauch04:_editin_descr_logic_ontol_with}). In initial experiments with ecologists and evolutionary biologists studying trait data, we found that \owlifier\ enabled them to quickly and easily comprehend and construct relatively complex and meaningful ontologies.

The rest of this paper is organized as follows. In \secref{sec:owlifier} we describe the basic syntax and semantics of \owlifier. The semantics of \owlifier\ blocks is given by mappings from \owlifier\ expressions to description-logic statements. Readers unfamiliar with description-logic notation may safely ignore these mappings, focusing instead on the descriptions and examples of blocks given in \secref{sec:owlifier}.  We define blocks that support a large subset of OWL-DL and that also generally follow the ontology creation guidelines defined in \citep{rector04:_owl_pizzas}. We also simplify certain aspects of ontology creation using OWL, e.g., by assuming classes are disjoint by default (unless specified otherwise) \citep{rector04:_owl_pizzas}. In \secref{sec:features} we describe additional characteristics of \owlifier\ and discuss issues with respect to classification and reasoning. In \secref{sec:implementation} we briefly describe the \owlifier\ implementation, and conclude in \secref{sec:conclusion} with related and future work. In general, the goal of \owlifier\ is not to support all constructs in OWL-DL, but instead to provide a higher-level ontology syntax (via spreadsheet blocks) that is easy for ecologists to use and understand while also providing the necessary constructs for developing typical ecological ontologies. By compiling \owlifier\ to OWL-DL, we also allow for experts to refine and extend the ontology using more advanced ontology editing tools if necessary (cf.~\figref{fig:owlifier}).

@@

\myblock{Import Blocks.} Import blocks assign namespace labels to
external ontologies. Each external ontology is imported into the
current ontology. We refer to the ontologies of import blocks as
\emph{imported ontologies}.  Using import blocks, classes and
properties of imported ontologies can be used within other blocks of
an \owlifier\ table.  Rows containing import blocks take the form
\begin{itemize}
\item[]
  \begin{tabular}{|l|l|l|}\hline
    \textsf{import} & $n$ & $u$ \\ \hline 
  \end{tabular} 
\end{itemize}
where $n$ is a namespace label and $u$ is an OWL ontology URI. Classes
and properties from imported ontologies are referenced by prefixing
the namespace label $n$ to the corresponding class or property name in
the normal way. As an example, the following block imports the SWEET
``Earth Realm'' ontology \citep{raskin05}
\begin{small}
\begin{tabbing}
  ~~\textsf{import}  \textsf{sweet} 
  \textsf{http://sweet.jpl.nasa.gov/ontology/earthrealm.owl}
\end{tabbing}
\end{small}
With this import block the class denoting Marine Ecosystems (a class
defined in the SWEET ontology) can be referred to from within an
\owlifier\ table using the expression
\textsf{sweet:MarineEcosystem}. Because this class refers to a class
in another ontology, we refer to it as an \emph{imported class}.

@@

\myblock{Relationship Blocks.} Relationship blocks define
\emph{required} class object properties.  An object property within
OWL is a property defined between two class instances. Rows containing
relationship blocks take the form
\begin{itemize}
\item[]
  \begin{tabular}{|l|l|l|l|l|l|}\hline \textsf{relationship} & $p$ & $c_1$
    & $c_2$ & \dots & $c_n$ \\ \hline
  \end{tabular} \hfill ($n \ge 2$)
\end{itemize}
where $p$ is an object property and each $c$ is a class.  For every
class $c_i$, the relationship block induces the description logic axiom $c_i
\sqsubseteq \exists p . c_{i+1}$ stating that each instance of $c_i$
is $p$-related to some instance of $c_{i+1}$, for $1 \le i < n$.  For
example, the following block states that instances of the class
California Voles live in Grassy Areas.
\begin{tabbing}
  ~~\textsf{relationship} \textsf{livesIn} \textsf{CaliforniaVole} 
    \textsf{GrassyArea}
\end{tabbing}
In some cases, a particular property can apply to a sequence of
classes, and for convenience, each such class can be specified in
\owlifier\ using a single block. For example, consider the following
block.
\begin{tabbing}
  ~~\textsf{relationship} \textsf{directlyBelow} \textsf{Hypolimnion} 
    \textsf{Thermocline} \textsf{Epilimnion} 
\end{tabbing}
This block states that, e.g., within a thermally stratified lake, the
Hypolimnion layer is directly below the Thermocline layer, and the
Thermocline layer is directly below the Epilimnion layer.

@@

\myblock{Sufficient Blocks.} Sufficient blocks are similar to synonym blocks in that they state equivalences between classes. We consider two types of sufficient blocks. A sufficient entity block takes the form
\begin{itemize}
\item[]
  \begin{tabular}{|l|l|l|l|l|l|}\hline \textsf{sufficient} & \textsf{entity} & $c_1$ & $c_2$ & \dots & $c_n$
\\ \hline
\end{tabular} \hfill ($n \ge 2$)
\end{itemize}
where each $c_i$ is a class. A sufficient entity block induces the description logic axiom $c_1 \equiv c_2 \sqcap \dots \sqcap c_n$ stating that the class $c_1$ is equivalent to the intersection of the classes $c_2$ through $c_n$.  Similarly, a sufficient relationship block takes the form
\begin{itemize}
\item[]
  \begin{tabular}{|l|l|l|l|l|l|l|l|}\hline \textsf{sufficient} & \textsf{relationship} & (\textsf{not}) & $p$ & $c_1$ & $c_2$ \\ \hline \end{tabular} \end{itemize} where $p$ is an object property, $c_1$ and $c_2$ are classes, and \textsf{not} is an optional qualifier.  A sufficient relationship block (without the \textsf{not} qualifier) induces the description logic axiom $c_1 \equiv \exists p . c_2$ stating that the class $c_1$ is equivalent to the set of individuals that are $p$-related to at least one individual of class $c_2$. The \textsf{not} qualifier states that the absence of the property is a defining characteristic of the class. Thus, a sufficient relationship block containing a \textsf{not} qualifier induce the description logic axiom $c_1 \equiv \neg \exists p . c_2$.  The use of multiple sufficient blocks for a particular class results in a single axiom in which each constraint (i.e., class constructor) is combined via intersection ($\sqcap$). For example, the following blocks 
\begin{small}
\begin{tabbing}
~~\textsf{sufficient} \= \textsf{relationship} \= \textsf{Eutheria} \= \textsf{Mammal} \= \kill 
~~\textsf{sufficient} \> \textsf{relationship} \> \textsf{hasPart} \> \textsf{Mammal} \> \textsf{Hair} \\
  ~~\textsf{sufficient} \> \textsf{entity} \> \textsf{Eutheria} \> \textsf{Mammal} \\
  ~~\textsf{sufficient} \> \textsf{relationship} \> \textsf{not} \> \textsf{hasPart} \> \textsf{Eutheria}  \textsf{EpipubicBone} 
\end{tabbing}
induce the following description logic axioms.
\begin{tabbing}
~~$\textsf{Mammal} \equiv \exists \textsf{hasPart}.\textsf{Hair}$\\ 
~~$\textsf{Eutheria} \equiv \textsf{Mamma} \sqcap \neg \exists \textsf{hasPart} . \textsf{EpipubicBone}$
\end{tabbing}
\end{small}
These axioms state that \emph{any} individual that has Hair as a part is a Mammal and that \emph{any} Mammal that does not have an Epipubic Bone as a part is a Eutheria.

@@

\figref{fig:owlifier} shows the general architecture of the \owlifier\ application. A scientist first creates a spreadsheet containing a set of ontology definitions using \owlifier\ conventions.  The spreadsheet is then exported as a plain-text file containing the \owlifier\ table, e.g., using a CSV or tab-delimited format. The \owlifier\ table is then sent to the \owlifier\ application, which performs a number of steps that include: (i) parsing the file; (ii) generating the corresponding OWL-DL representation (e.g., in the current implementation, the OWL-API is used \citep{horridge07:_ignit_owl}, although Jena \citep{carroll04:_jena} could be used as well); (iii) validating the ontology and ensuring it is consistent via an OWL-DL reasoner (e.g., the current implementation uses Pellet \citep{sirin07:_pellet}); and (iv) outputting the corresponding OWL-DL file. The resulting OWL file can then be refined and extended via existing OWL tools (e.g., Protege or SWOOP). It is also possible for the refined ontology to be converted back to a corresponding \owlifier\ representation (shown using dashed lines in \figref{fig:owlifier}). While straightforward to perform, this ``back'' conversion is not information preserving since not all OWL-DL statements can be represented through \owlifier\ blocks.  That is, the conversion from OWL-DL to \owlifier\ will only preserve OWL-DL constructs that are represented by \owlifier\ blocks.  In particular, \owlifier\ does not support a number of OWL-DL language constructs, which include domain and range constraints, datatype properties, sub properties, property constraints other than transitive (i.e., functional, symmetric, inverse, inverse functional), value restrictions, individuals, etc., as well as various combinations of OWL-DL constructs.

@@

This paper presents a new approach for developing ontologies to address barriers in ontology development and adoption by ecologists. Our approach allows scientists to use familiar spreadsheet software (e.g., frequently used by ecologists for storing and analyzing data) to create ontologies by filling in a set of templates, or blocks, that generate OWL-DL class hierarchies, properties, and constraints via the \owlifier\ application.  This approach can provide a more intuitive and accessible ontology editing environment for ecologists, especially compared to existing OWL-based tools that require considerable expertise in the underlying logic formalisms. Similar to standard OWL ontologies, \owlifier\ spreadsheets can import and extend existing OWL-DL ontologies (including those generated from \owlifier\ spreadsheets), which further supports the community development and interoperability of ontologies. Further, \owlifier\ blocks (or similarly, partially filled in blocks) can be reused during ontology creation to help ensure that future ontology development adheres to best practices and community standards (e.g., similar to ``part-of'' properties defined in the Gene Ontology \citep{bada04:_short_study_succes_of_gene_ontol}).

@@

As future work, we plan to extend our current \owlifier\ implementation to support the full set of blocks defined here as well as introduce additional blocks, e.g., for creating OWL datatype properties. We also would like to explore approaches for supporting round-trip conversions between \owlifier\ tables and OWL-DL ontologies that allow scientists and knowledge-representation experts to incrementally develop \owlifier-based ontologies.  Specifically, we want to extend the \owlifier\ application so that it can re-apply changes made by knowledge-representation experts that previously modified the OWL-DL ontology generated by \owlifier. For instance, if an OWL-DL ontology generated from an \owlifier\ spreadsheet is refined and extended by a knowledge-representation expert, then converted back into an \owlifier\ ontology that is further edited and extended by a scientist, and then converted again into an OWL-DL ontology, we want to maintain (i.e., re-apply) the original extensions and edits created by the knowledge representation expert that are still relevant (thus supporting lossless conversions).  We also plan to extend the \owlifier\ application to support translation into the observation ontology framework presented in \citep{bowers08:_concep_model_framew_for_expres} and perform additional testing and evaluation of the \owlifier\ approach with a wide range of ecologists.

>>>>

Efforts to engage scientists in the development of ontologies
typically leverage the W3C Web Ontology Language (OWL)
\citep{smith04:_owl_web_ontol_languag_guide} as a standard XML syntax
for representing and sharing ontologies. A key advantage of OWL is
that it is supported by a wide range of generic tools, including
editors
\citep{knublauch04:_editin_descr_logic_ontol_with,kalyanpur05:_swoop},
reasoning systems
\citep{sirin07:_pellet,tsarkov06:_fact_descr_logic_reason}, query
languages
\citep{prudhommeaux08:_sparq_query_languag_for_rdf,motik05:_query_answer_for_owl_dl_with_rules},
and storage technologies
\citep{carroll04:_jena,broekstra02:_sesam}. However, most of these
tools are primarily targeted at experts in knowledge engineering and
software development familiar with the underlying description logic
semantics of OWL-DL \citep{grau08:_owl}. This is especially true with
ontology editors (such as Protege, SWOOP, etc.), which allow for very
detailed ontology specifications, but at the same time require
considerable understanding of the underlying ontology formalisms and
syntax.  The development of the Gene Ontology (GO) has been underway
for over a decade within the molecular biology community, and has lead
to significant improvements in data interoperability by enriching
genomic and biomedical data resources with annotations to
community-based ontologies. A major lesson learned from these efforts
was the importance of keeping the ontology focused, and indeed GO has
a highly delimited scope
\citep{bada04:_short_study_succes_of_gene_ontol}. Ecology, on the
other hand, encompasses an extremely broad range of scales and
disciplines, such that a coordinated ontology development effort would
need to involve a large number of experts to cover the diverse types
of information relevant to ecological investigations.  We believe that
ecological researchers must also begin constructing ontologies for
their own specialized subdisciplines. However, due to the substantial
expertise in formal knowledge representation that is required to use
most existing ontology editing tools, new tools and approaches are
needed to enable the wide-scale creation and adoption of ontologies in
ecology.

This paper presents a novel approach for ontology creation that aims
at being more intuitive for ecologists and that can be used to rapidly
construct large ontologies of ecological terms (classes and
properties). These terms can then be used in subsequent steps to
annotate ecological data (instances), e.g., as in
\citep{bowers08:_concep_model_framew_for_expres,berkley09:_improv_data_discov_metad_repos_seman_searc}.
Our approach is to allow scientists to use common and familiar
spreadsheet-based tools (e.g., Microsoft Excel, Apple Numbers, Open
Office Spreadsheet) to describe, in an intuitive way, different
aspects of an ontology, and then to take these descriptions
(represented as spreadsheets) and convert them into full-fledged OWL
ontologies using a software application called \owlifier.

An \owlifier\ spreadsheet consists of a set of \emph{blocks} that have
a predefined template structure for users to fill in. Each non-empty
row in an \owlifier\ spreadsheet constitutes a block. Each block
defines different aspects of an ontology including classes,
subclasses, synonyms, and properties.  We also provide blocks for
plain-text descriptions of classes and properties, and for referencing
one or more existing ontologies (e.g., to extend an existing ontology
or to define ontology articulations). Blocks can be sparse (inheriting
from previous blocks), which further simplifies the creation of large
ontologies by minimizing the amount of redundant information that must
be provided in the \owlifier\ tabular representation. Several of the
more advanced OWL features are omitted in \owlifier, primarily
associated with properties (e.g., role inclusion axioms, domain and
range property restrictions, and data-type properties among others),
as these can be confusing for non-experts and are more suitable for
experienced knowledge modelers.

While not as expressive as OWL-DL, \owlifier\ can be used to produce
ontology structures that are essential for improved data discovery and
integration \citep{madin07:_ontol_for_descr_and_synth}. Just as
important, because spreadsheet tools are already frequently used by
ecologists to store and analyze data, \owlifier\ can provide
ecologists with a familiar and accessible user interface for ontology
creation. This approach also leverages the easy-to-use interfaces
provided by many spreadsheet tools for organizing and manipulating
tabular data, e.g., via cut/copy/paste, search/replace, sort,
split-window, track-changes, freeze panes, and so on.  In this way, an
ecologist can easily construct (or load) a set of terms, and then
incrementally organize these into class hierarchies, properties, and
constraints.  The use of an editing environment that is familiar to
scientists can significantly help improve the speed and understanding
of ontology construction and avoid the often time-consuming task of
locating, downloading, installing, and learning fundamentally new
software applications and interfaces (e.g., CMAP-COE
\citep{hayes05:_collab_knowl_captur_in_ontol}, Protege
\citep{knublauch04:_editin_descr_logic_ontol_with}). In initial
experiments with ecologists and evolutionary biologists studying trait
data, we found that \owlifier\ enabled them to quickly and easily
comprehend and construct relatively complex and meaningful ontologies.

The rest of this paper is organized as follows. In
\secref{sec:owlifier} we describe the basic syntax and semantics of
\owlifier. The semantics of \owlifier\ blocks is given by mapping
\owlifier\ expressions to description-logic statements. Readers
unfamiliar with description-logic notation may safely ignore these
mappings, focusing instead on the descriptions and examples of blocks
given in \secref{sec:owlifier}.  We define blocks that support a large
subset of OWL-DL and that also generally follow the ontology creation
guidelines defined in \citep{rector04:_owl_pizzas}. We also simplify
certain aspects of ontology creation using OWL, e.g., by assuming
classes are disjoint by default (unless specified otherwise)
\citep{rector04:_owl_pizzas}. In \secref{sec:features} we describe
additional characteristics of \owlifier\ and discuss issues with
respect to classification and reasoning. In
\secref{sec:implementation} we briefly describe the
\owlifier\ implementation, and conclude in \secref{sec:conclusion}
with related and future work. In general, the goal of \owlifier\ is
not to support all constructs in OWL-DL, but instead to provide a
higher-level ontology syntax (via spreadsheet blocks) that is easy for
ecologists to use and understand while also providing the necessary
constructs for developing typical ecological ontologies. By compiling
\owlifier\ to OWL-DL, we also allow for experts to refine and extend
the ontology using more advanced ontology editing tools if necessary
(cf.~\figref{fig:owlifier}).

@@

\myblock{Import Blocks.} Import blocks assign namespace labels to
external ontologies. Each external ontology is imported into the
current ontology. We refer to the ontologies of import blocks as
\emph{imported ontologies}.  Using import blocks, classes and
properties of imported ontologies can be used within other blocks of
an \owlifier\ table.  Rows containing import blocks take the form
\begin{itemize}
\item[]
  \begin{tabular}{|l|l|l|}\hline
    \textsf{import} & $n$ & $u$ \\ \hline 
  \end{tabular} 
\end{itemize}
where $n$ is a namespace label and $u$ is an OWL ontology URI. Classes
and properties from imported ontologies are referenced by prefixing
the namespace label $n$ to the corresponding class or property name in
the normal way. As an example, the following block imports the SWEET
``Earth Realm'' ontology \citep{raskin05}.
\begin{small}
\begin{tabbing}
  ~~\textsf{import}  \textsf{sweet} 
  \textsf{http://sweet.jpl.nasa.gov/ontology/earthrealm.owl}
\end{tabbing}
\end{small}
With this import block the class denoting Marine Ecosystems (a class
defined in the SWEET ontology) can be referred to from within an
\owlifier\ table using the expression
\textsf{sweet:MarineEcosystem}. Because this class refers to a class
in another ontology, we refer to it as an \emph{imported class}.

@@

\myblock{Relationship Blocks.} Relationship blocks define
\emph{required} class object properties.  An object property within
OWL is a property defined between two class instances. Rows containing
relationship blocks take the form
\begin{itemize}
\item[]
  \begin{tabular}{|l|l|l|l|l|l|}\hline \textsf{relationship} & $p$ & $c_1$
    & $c_2$ & \dots & $c_n$ \\ \hline
  \end{tabular} \hfill ($n \ge 2$)
\end{itemize}
where $p$ is an object property and each $c$ is a class.  For every
class $c_i$, the relationship block induces the description logic axiom $c_i
\sqsubseteq \exists p . c_{i+1}$ stating that each instance of $c_i$
is $p$-related to some instance of $c_{i+1}$, for $1 \le i < n$.  For
example, the following block states that instances of the class
California Voles live in Grassy Areas.
\begin{tabbing}
  ~~\textsf{relationship} \textsf{livesIn} \textsf{CaliforniaVole} 
    \textsf{GrassyArea}
\end{tabbing}
In some cases, a particular property can apply to a sequence of
classes, and for convenience, each such class can be specified in
\owlifier\ using a single block. For example, consider the following
block.
\begin{small}
\begin{tabbing}
  ~~\textsf{relationship} \textsf{directlyBelow} \textsf{Hypolimnion} 
    \textsf{Thermocline} \textsf{Epilimnion} 
\end{tabbing}
\end{small}
This block states that, e.g., within a thermally stratified lake, the
Hypolimnion layer is directly below the Thermocline layer, and the
Thermocline layer is directly below the Epilimnion layer.

@@

\myblock{Sufficient Blocks.} Sufficient blocks are similar to synonym
blocks in that they state equivalences between classes. We consider
two types of sufficient blocks. A sufficient entity block takes the
form
\begin{itemize}
\item[]
  \begin{tabular}{|l|l|l|l|l|l|}\hline \textsf{sufficient} & \textsf{entity} & $c_1$ & $c_2$ & \dots & $c_n$
\\ \hline
\end{tabular} \hfill ($n \ge 2$)
\end{itemize}
where each $c_i$ is a class. A sufficient entity block induces the
description logic axiom $c_1 \equiv c_2 \sqcap \dots \sqcap c_n$
stating that the class $c_1$ is equivalent to the intersection of the
classes $c_2$ through $c_n$.  Similarly, a sufficient relationship
block takes the form
\begin{itemize}
\item[]
  \begin{tabular}{|l|l|l|l|l|l|l|l|}\hline \textsf{sufficient} & \textsf{relationship} & (\textsf{not}) & $p$ & $c_1$ & $c_2$ \\ 
\hline \end{tabular} 
\end{itemize} 
where $p$ is an object property, $c_1$ and $c_2$ are classes, and
\textsf{not} is an optional qualifier.  A sufficient relationship
block (without a \textsf{not} qualifier) induces the description logic
axiom $c_1 \equiv \exists p . c_2$ stating that the class $c_1$ is
equivalent to the set of individuals that are $p$-related to at least
one individual of class $c_2$. The \textsf{not} qualifier states that
the absence of the property is a defining characteristic of the
class. Thus, a sufficient relationship block containing a \textsf{not}
qualifier induces the description logic axiom $c_1 \equiv \neg \exists
p . c_2$.  The use of multiple sufficient blocks for a particular
class results in a single axiom in which each constraint (i.e., class
constructor) is combined via intersection ($\sqcap$). For example, the
following blocks
\begin{small}
\begin{tabbing}
~~\textsf{sufficient} \= \textsf{relationship} \= \textsf{Eutheria} \=
  \textsf{Mammal} \= \kill ~~\textsf{sufficient} \>
  \textsf{relationship} \> \textsf{hasPart} \> \textsf{Mammal} \>
  \textsf{Hair} \\ ~~\textsf{sufficient} \> \textsf{entity} \>
  \textsf{Eutheria} \> \textsf{Mammal} \\ ~~\textsf{sufficient} \>
  \textsf{relationship} \> \textsf{not} \> \textsf{hasPart} \>
  \textsf{Eutheria} \textsf{EpipubicBone}
\end{tabbing}
induce the following description logic axioms.
\begin{tabbing}
~~$\textsf{Mammal} \equiv \exists \textsf{hasPart}.\textsf{Hair}$\\ 
~~$\textsf{Eutheria} \equiv \textsf{Mamma} \sqcap \neg \exists \textsf{hasPart} . \textsf{EpipubicBone}$
\end{tabbing}
\end{small}
These axioms state that \emph{any} individual that has Hair as a part
is a Mammal and that \emph{any} Mammal that does not have an Epipubic
Bone as a part is a Eutheria.

@@

\figref{fig:owlifier} shows the general architecture of the
\owlifier\ application. A scientist first creates a spreadsheet
containing a set of ontology definitions using \owlifier\ conventions.
The spreadsheet is then exported as a plain-text file containing the
\owlifier\ table, e.g., using a CSV or tab-delimited format. The
\owlifier\ table is then sent to the \owlifier\ application, which
performs a number of steps that include: (i) parsing the file; (ii)
generating the corresponding OWL-DL representation (e.g., in the
current implementation, the OWL-API is used
\citep{horridge07:_ignit_owl}, although Jena \citep{carroll04:_jena}
could be used as well); (iii) validating the ontology and ensuring it
is consistent via an OWL-DL reasoner (e.g., the current implementation
uses Pellet \citep{sirin07:_pellet}); and (iv) outputting the
corresponding OWL-DL file. The resulting OWL file can then be refined
and extended via existing OWL tools (e.g., Protege or SWOOP). It is
also possible for the refined ontology to be converted back to a
corresponding \owlifier\ representation (shown using dashed lines in
\figref{fig:owlifier}). While straightforward to perform, this
``back'' conversion is not information preserving since not all OWL-DL
statements can be represented through \owlifier\ blocks.  That is, the
conversion from OWL-DL to \owlifier\ will only preserve OWL-DL
constructs that are represented by \owlifier\ blocks.  In particular,
\owlifier\ does not support a number of OWL-DL language constructs,
including domain and range constraints, datatype properties, sub
properties, property constraints (i.e., functional, symmetric, and
inverse functional), value restrictions, individuals, etc., as well as
various combinations of OWL-DL constructs.

@@

This paper presents a new approach for developing ontologies to
address barriers in ontology development and adoption by
ecologists. Our approach allows scientists to use familiar spreadsheet
software (e.g., frequently used by ecologists for storing and
analyzing data) to create ontologies by filling in a set of templates,
or blocks, that generate OWL-DL class hierarchies, properties, and
constraints via the \owlifier\ application.  This approach can provide
a more intuitive and accessible ontology editing environment for
ecologists, especially compared to existing OWL-based tools that
require considerable expertise in the underlying logic
formalisms. Similar to standard OWL ontologies,
\owlifier\ spreadsheets can import and extend existing OWL-DL
ontologies (including those generated from \owlifier\ spreadsheets),
which further supports the community development and interoperability
of ontologies. Blocks in \owlifier\ (or similarly, partially filled in
blocks) can also be reused during ontology creation to help ensure
that future ontology development adheres to best practices and
community standards (e.g., similar to ``part-of'' properties defined
in the Gene Ontology
\citep{bada04:_short_study_succes_of_gene_ontol}).

@@

As future work, we plan to extend our current
\owlifier\ implementation to support the full set of blocks defined
here as well as introduce additional blocks, e.g., for creating OWL
datatype properties. We also would like to explore approaches for
supporting round-trip conversions between \owlifier\ tables and OWL-DL
ontologies that allow scientists and knowledge-representation experts
to incrementally develop \owlifier-based ontologies.  Specifically, we
want to extend the \owlifier\ application so that it can store and
re-apply changes made by knowledge-representation experts that
previously modified an OWL-DL ontology generated by \owlifier. For
instance, if an OWL-DL ontology generated from an
\owlifier\ spreadsheet is refined and extended by a
knowledge-representation expert, then converted back into an
\owlifier\ ontology that is further edited and extended by a
scientist, and then converted again into an OWL-DL ontology, we want
to maintain (i.e., re-apply) the original extensions and edits created
by the knowledge-representation expert that are still relevant (thus
supporting lossless conversions).  We also plan to extend the
\owlifier\ application to support translation into the observation
ontology framework presented in
\citep{bowers08:_concep_model_framew_for_expres} and perform
additional testing and evaluation of the \owlifier\ approach with a
wide range of ecologists.
